--- 
title: "Practical Psychometrics"
subtitle: "A Psychological Assessment Toolkit"
author: "W. Joel Schneider"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output:
  bookdown::tufte_html_book:
    toc: TRUE
    toc_depth: 2
    css: [toc.css,equity.css,mycss.css]
    tufte_features: []
    keep_md: true
    split_by: chapter
    highlight: pygments
    includes: 
      in_header: mj_config.html
documentclass: book
bibliography: [book.bib, packages.bib]
csl: apa.csl
link-citations: yes
github-repo: wjschne/toolkit
nocite: |
  @R-bookdown, @R-dplyr, @R-extrafont, @R-fBasics, @R-forcats, @R-fMultivar, @R-gganimate, @R-ggforce, @R-ggplot2, @R-IDPmisc, @R-knitr, @R-knitr, @R-purrr, @R-rmarkdown, @R-sjmisc, @R-sn, @R-stringr, @R-base, @R-tibble, @R-tidyr, @R-tidyverse, @R-tikzDevice, @R-tufte
---

# Preface {-}

I have ambitious goals for this book, but it is not nearly complete. I have been working on it off and on since 2012. Eventually it will have one or more R packages to accompany it. 

```{r setup, include=FALSE}

# source("loader.R")

packages <- c(
  .packages())
knitr::write_bib(packages, 'packages.bib')
```

```{r makesvg, eval=F}
c("SUDS",
  "DiscreteContinuous",
  "Stevens",
  "dichotomy",
  "Likert",
  "dice",
  "pdfIllustration",
  "pmf",
  "cdfDie",
  "QuantileFunction",
  "coin",
  "twocoin",
  "halfnormal") %>% walk(pdf2svg)
```

Most of the figures and tables for this book were created in R or in $\LaTeX$. To make the content as accessible and transparent as possible, I have included buttons that will reveal the code used to make each figure or table. For example, the initial setup code used for this book can be seen by clicking the button below:

```{r setupCode, code = readLines("loader.R"), button_r = TRUE}

```

In addition, all the files and code used to create this book can be found in its [Github repository](https://github.com/wjschne/toolkit).




